---
title: "Mini Lab: T Tests in R"
editor: visual
format: html
execute: 
  echo: true
  eval: false
---

```{r}
#| echo: false
#| eval: true
knitr::opts_chunk$set(comment = NA)

```

In this short assignment, you will practice hypothesis testing in R. The graded portion of the assignment will only be to answer the questions in the Canvas quiz.

## Load Data

Most of the time we load data from csv files we have downloaded on our computer. Sometimes, R or R packages will have datasets included, so you do not need to load the data from a file, but instead can use the `data()` function. We will use a built-in dataset that includes arrest data for each US state in 1973.

Load the data with this code:

```{r}
#| eval: true

data("USArrests")
```

If the data does not immediately show up in your environment, it might just need a little more time, or you might need to call the data first with a function. Try `head(USArrests)`.

## Hypothesis testing

Suppose I heard that in 2021, the arrest rate for assault was 135 per 100,000. I want to see if the average state-level rate 1973 was significantly different from that at the 99% confidence level. a. Write down the null and alternative hypotheses. b. Use the following function to do the hypothesis test:

```{r}
t.test(mydata, alternative, mu, conf.level)
```

Where "mydata" is your data; "alternative" can take on one of three values: "two.sided", "greater", "less"; mu is $\mu$; and conf.level is $(1-\alpha)$. As an example, check out this hypothesis test for stock data:

```{r}
t.test(stock, alternative = "two.sided", mu = 170, conf.level = 0.95)
```

c.  Write down your conclusion.

## Creating a dummy variable

I want to break the data into two groups: states with a high urban population, and states with a low urban population. `UrbanPop` is a variable that says the percent of the state which is urban. If I run the code:

```{r}
summary(USArrests$UrbanPop)
```

I see that the median percent urban is 66. I will choose that as my cutoff for urban versus non-urban. To create a variable that equals 1 if the state is urban, and 0 if it's not, I will use the code:

```{r}
USArrests$urban <- ifelse(USArrests$UrbanPop > 66, 1, 0)
```

## Two sample t tests

In the examples we've done so far, we test our sample mean against the mean for a larger population or theoretical distribution. Sometimes, we want to test whether two groups are different based on two different samples. In this case, I want to know if murder rates are higher in "urban" states, so a two-sample t-test is appropriate.

Hypothesis testing for two samples is slightly different, and the calculation for t is slightly different.

a.  Write down hypotheses. $$ H_0: \mu_{urban} \leq \mu_{non-urban} $$ $$ H_1: \mu_{urban} > \mu_{non-urban} $$ Which is equivalent to: $$ H_0: \mu_{urban} - \mu_{non-urban} \leq 0$$ $$ H_1: \mu_{urban} - \mu_{non-urban} > 0 $$

b.  Perform the t test. If I were to be calculating this by hand, the formula would be: $$ t= \frac{(\bar x_{urban}-\bar x_{non-urban})-(\mu_{urban}-\mu_{non-urban})}{{\sqrt{\frac{s_{urban}^2}{n_{urban}} + \frac{s_{non-urban}^2}{n_{non-urban}} }}}$$ If you are interested, you can calculate this on your own. The degrees of freedom would be $df = n_{urban} + n_{non-urban}-2$.

I could calculate the mean, standard deviation, and sample size for each group like so:

```{r}
mean(USArrests$Murder[USArrests$urban == 1])
sd(USArrests$Murder[USArrests$urban == 1])
length(USArrests$Murder[USArrests$urban == 1])
```

However, we can use the t test formula in R as well. The formula for two-sample t tests is:

```{r}
t.test(data1, data2, alternative, conf.level)
```

The data you want to compare are different rows in the same dataframe. You will have to subset your data, either within the formula using brackets, or by making new data frames using brackets or `dplyr`. c. Write down your conclusions.
